$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-06 paper/paper.tex
Shorten swapping and subject congruence

\subsection{Notation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\autoref{fig:notation} illustrates the notation used in this paper.
\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\type{\N} : \Set}
    \; \textsc{Nat}

@@

Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
We otherwise use constructor names ($\constr{0}$ and $\suc$) to name our typing rules.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted by constructors.

@@

\begin{remark}
  It is possible for the usage annotations of a variable $i$ within a process $P$ to be preserved as leftovers, yet for the variable $i$ to be mentioned by $P$: $P$ can choose to send $\lz$ usage annotations of $i$ over a channel.
\end{remark}

\todo{sw and swap? Just use one}

Swapping two variables preserves the well-typedness of a process.
Let $\func{sw}_i$ be a function that swaps variables $i$ and $\suc i$.
For simplicity, we overload $\func{sw}$ and use it to swap types in typing contexts and usage annotations in usage contexts. We use $\func{swap}_i \; P$ to swap variables $i$ and $\suc i$ within a process $P$, as introduced in structural congruence (\autoref{structural-congruence}).
\begin{nitheorem}[Swapping]
  \label{thm:swapping}
  Let $P$ be a well-typed process in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{sw}_i \; \gamma}{\func{sw}_i \; \Gamma}{\func{swap}_i \; P}{\func{sw}_i \; \Xi}$.
\end{nitheorem}

\begin{remark}
  We limit ourselves to the swapping of adjacent variables $i$ and $\suc i$.
  Doing so is enough to prove subject reduction.
  More generally, any two variables can be swapped by a series of adjacent variable swappings.
\end{remark}

@@

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we will only consider the base cases for $\constr{struct}$ (and their symmetric variants under $\constr{sym}$), which are nontrivial. Full proof in \cite{Zalakain2020Agda}.
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
      %The point at which $Q$ and $P$ join in $\comp{Q}{P}$ will no longer be the same.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$..
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we use \autoref{lm:types-unused} to show that $P$ preserves the usage annotations of the unused variable.
      We then use strengthening (\autoref{thm:strengthening}) to delete the unused variable from $P$, and compose the result with the typing derivation on $Q$ wrapped in $\constr{chan}$.
      In the reverse direction, we weaken (\autoref{thm:weakening}) $P$ so that it accepts the additional variable.
      We then use \autoref{lm:lower-lift} to show that lowering and then lifting $P$ results in $P$.
    \item
      Case $\constr{scope-comm}$:
      we use swapping (\autoref{thm:swapping}) (and \autoref{lm:swap-swap} for the reverse direction) to show that swapping two elements in $P$ twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}
\todo{remove the space between the itemize and the qed at the end}
\subsection{Substitution}
\label{substitution}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{generaliz* to generalis* everywhere}
\autoref{thm:subst-generalization} proves a generalized version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalized version to the concrete case where $i$ is $\constr{0}$, as required by subject reduction.
\todo{explain what i is 0 mean}

>>>>

\subsection{Notation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\type{\N} : \Set}
    \; \textsc{Nat}

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
We otherwise use constructor names ($\constr{0}$ and $\suc$) to name our typing rules.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted by constructors.

@@

\begin{remark}
  Usage annotations of a variable $i$ within a process $P$ can be preserved as leftovers, yet the variable $i$ be mentioned by $P$: $P$ may send $\lz$ usage annotations of $i$ over a channel.
\end{remark}

Swapping two variables preserves the well-typedness of a process.
For simplicity, we extend $\func{swap}_i$ introduced in \autoref{structural-congruence} to swap types in typing contexts and usage annotations in usage contexts.
\begin{nitheorem}[Swapping]
  \label{thm:swapping}
  Let $P$ be a well-typed process in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{swap}_i \; \gamma}{\func{swap}_i \; \Gamma}{\func{swap}_i \; P}{\func{swap}_i \; \Xi}$.
\end{nitheorem}

@@

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we only consider the nontrivial base cases for $\constr{struct}$ and their symmetric variants.
  Full proof in \cite{Zalakain2020Agda}.
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$..
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we to show that $P$ preserves the usage annotations of the unused variable (\autoref{lm:types-unused}) and then use strengthening (\autoref{thm:strengthening}).
      In the reverse direction, we use weakening (\autoref{thm:weakening}) on $P$ and show that lowering and then lifting $P$ results in $P$ (\autoref{lm:lower-lift}).
    \item
      Case $\constr{scope-comm}$:
      we use swapping (\autoref{thm:swapping}), and for the reverse direction swapping and \autoref{lm:swap-swap} to show that swapping two elements in $P$ twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}
\todo{remove the space between the itemize and the qed at the end}

\subsection{Substitution}
\label{substitution}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{generaliz* to generalis* everywhere}
\autoref{thm:subst-generalization} proves a generalized version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalized version to the concrete case where $i$ is $\constr{0}$, as required by subject reduction.
\todo{explain what i is 0 mean}
